---
title: "R Notebook"
output: html_notebook
---
Following the script of: Delgadillo-Ordoñez, N., Garcias-Bonet, N., Raimundo, I. et al. Probiotics reshape the coral microbiome in situ without detectable off-target effects in the surrounding environment. Commun Biol 7, 434 (2024). https://doi.org/10.1038/s42003-024-06135-3

# Libraries
```{r}
library(tidyverse) #Collection of packages designed for data science.
library(ggplot2) #Package to create graphs (see this link for cheatsheet: https://ggplot2.tidyverse.org)
library(vegan) #Comprehensive toolbox for ecological data analysis, particularly for multivariate data.
library(pairwiseAdonis) #is an extension of the adonis function from the vegan package, providing pairwise comparisons 
#for multivariate analysis of variance using distance matrices. 
#It is particularly useful for identifying significant differences between groups in complex datasets.
library(dplyr) #Provides grammar for data manipulation.
library(scales) #Provides tools to customize graphs.
library(grid) #Allows fine tuning of graphs, often used to enhance the aspect of ggplot2 graphs.
library(phyloseq) #is designed for the analysis of microbiome census data. It provides tools 
#for importing, storing, analyzing, and visualizing microbiome data.
library(grafify) #is designed to help with creating and customizing publication-ready graphics.
#It provides functions to enhance ggplot2 plots
#install.packages(microbiome)library(microbiome) #Provides toolpackas for analyzing and visualizing microbiome data.
library(dunn.test) #Provides the Dunn's test for multiple comparisons following a Kruskal-Wallis test. 
#This test is useful for non-parametric post-hoc analysis to determine which groups differ from each other.
library(stats) #Core package in R that provides a wide range of statistical functions and tools.
library(ggordiplots) #used to visualize  ordination results (e.g., PCA, NMDS, PCoA) using ggplot2. 
#It provides functions to create plots that are useful for interpreting and presenting multivariate analysis results.
library(DescTools) #Provides a range of functions for descriptive statistics, data transformation, and statistical tests.
library(ANCOMBC) #provides tools for performing the Analysis of Composition of Microbiomes with Bias Correction (ANCOM-BC). 
#It is designed for differential abundance analysis in microbiome data, specifically addressing biases and improving the accuracy of statistical inference.
library(psadd)
library(MiscMetabar)
library(stringr)
library(readxl) 
library(patchwork)
library(viridis)
library(colorBlindness)
library(ggpubr)
library(microbiome)
```

# Data input
Import rawdata and repair mapfile (metdata)
```{r}
taxfile_tourist <- read.csv( header=TRUE, row.names=1) #import taxonomy file (put path before header)
mapfile_tourist <- read_excel() #imports metadata file as excel
mapfile_tourist <- as.data.frame(mapfile_tourist) 
rownames(mapfile_tourist) <- mapfile_tourist[[1]] # asign rownames 
mapfile_tourist[[1]] <- NULL  # remove the now-duplicated column #delete column one which is now rownames
OTUs_tourist <- read.csv( header=TRUE, row.names=1) #import OTU file
```

# Data prep
clean up OTU file
```{r}
OTUs_tourist <- OTUs_tourist[,- 1] #delete first column that gives numbers of OTUs
OTUs_tourist <- OTUs_tourist[!rownames(OTUs_tourist) %in% c("P1_mock", "P2_mock", "P3_mock"), ] #delete the rows with the control samples
OTUs_tourist_transposed <- t(OTUs_tourist) #switch rows and columns so that OTU and taxonomy file are in the same format
```

Tables sanity check
```{r}
class(mapfile_tourist$Year) 
class(mapfile_tourist$Status)
as.character(mapfile_tourist$Year)
```

Clean up mapfile and correct spelling mistakes 
```{r}
mapfile_tourist <- mapfile_tourist[!rownames(mapfile_tourist) %in% c("CU23F_CI", "CUB:EK20_TRST_001", "CU21_TRST_030"), ] #delete rows that don't contain data
mapfile_tourist <- mapfile_tourist[!(mapfile_tourist$Site %in% c("Jan Kok")), ]
## fix spelling mistakes throughout columns in metadata file (mapfile)
fix_spelling <- c("URBNA" = "URBAN",
                  "TOURST" = "TOURIST",
                  "CONTROL1-2DB 2022" = "CONTROL1-2DB2022",
                  "CONTROL2-3PJ 2020" = "CONTROL2-3PJ2020",
                  "Porto Marie" = "Porto_Marie",
                  "Coral Estate" = "Coral_Estate",
                  "Playa Jeremy" = "Playa_Jeremy",
                  "Cas Abou" = "Cas_Abou",
                  "Coral Estate" = "Coral_Estate",
                  "Grote Knip" = "Grote_Knip",
                  "Mambo Beach" = "Mambo_Beach",
                  "Holiday Beach" = "Holiday_Beach",
                  "Directors Bay" = "Directors_Bay")

# Apply to all columns that contain the misspelled words
mapfile_tourist <- mapfile_tourist %>%
  mutate(across(
    c( Label, Status, Site),
    ~ str_replace_all(as.character(.), fix_spelling)
  ))
```

transform dataframes
```{r}
## Transform taxfile in a matrix
taxfile_tourist <- as.matrix(taxfile_tourist, nrow=nrow(taxfile_tourist))

class(taxfile_tourist)
head(taxfile_tourist)

## Transform mapfile (metadata file) in Dataframe
mapfile_tourist <- data.frame(mapfile_tourist, nrow=nrow(mapfile_tourist))
class(mapfile_tourist)

mapfile_tourist$Year_Site <- paste(mapfile_tourist$Year, mapfile_tourist$Site, sep = "_")

list(mapfile_tourist$Year_Site)
unique(mapfile_tourist$Year_Site)
list(mapfile_tourist$ID)

##Transform OTUs_counts into a matrix 

OTUs_tourist_transposed <- as.matrix(OTUs_tourist_transposed, nrow=nrow(OTUs_tourist_transposed))

class(OTUs_tourist_transposed)
```

Reoder mapfile 
```{r}
## Order variable levels (by default is alphabetic order, you want your control to be the reference) 
mapfile_tourist$Year_Site <- ordered(mapfile_tourist$Year_Site, levels= c("2020_Directors_Bay", "2020_Playa_Jeremy", "2020_Holiday_Beach", "2020_Coral_Estate", "2020_Cas_Abou", "2020_Mambo_Beach", "2020_Porto_Marie", "2020_Grote_Knip", "2021_Directors_Bay", "2021_Playa_Jeremy", "2021_Holiday_Beach", "2021_Coral_Estate", "2021_Cas_Abou", "2021_Mambo_Beach", "2021_Porto_Marie", "2021_Grote_Knip", "2022_Directors_Bay", "2022_Playa_Jeremy", "2022_Holiday_Beach", "2022_Coral_Estate", "2022_Cas_Abou", "2022_Mambo_Beach", "2022_Porto_Marie", "2022_Grote_Knip", "2023_Directors_Bay", "2023_Playa_Jeremy", "2023_Holiday_Beach", "2023_Coral_Estate", "2023_Cas_Abou", "2023_Mambo_Beach", "2023_Porto_Marie", "2023_Grote_Knip"))

levels(mapfile_tourist$Year_Site)

mapfile_tourist$Year <- ordered (mapfile_tourist$Year, levels = c("2020", "2021", "2022", "2023"))
levels(mapfile_tourist$Year)

mapfile_tourist$Site <- ordered (mapfile_tourist$Site, levels = c("Directors_Bay","Playa_Jeremy", "Holiday_Beach","Coral_Estate", "Cas_Abou", "Mambo_Beach", "Porto_Marie","Grote_Knip"))
levels(mapfile_tourist$Site)
```

# Phyloseq
Create combined phyloseq and clean it up
```{r}
phyloseq_tourist <- phyloseq(otu_table(OTUs_tourist_transposed, taxa_are_rows=TRUE), tax_table(taxfile_tourist), sample_data(mapfile_tourist))
phyloseq_tourist


#phyloseq-class experiment-level object
#otu_table()   OTU Table:         [ 2403 taxa and 135 samples ]
#sample_data() Sample Data:       [ 135 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 2403 taxa by 7 taxonomic ranks ]

# Filter out chloroplasts and mitochondria 
phyloseq_tourist_clean <- phyloseq_tourist %>%
  subset_taxa(
    Kingdom == "Bacteria" &
      Family  != "Mitochondria" &
      Order   != "Chloroplast"
    
  )
phyloseq_tourist_clean

#phyloseq-class experiment-level object
#otu_table()   OTU Table:         [ 2344 taxa and 135 samples ]
#sample_data() Sample Data:       [ 135 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 2344 taxa by 7 taxonomic ranks ]


phyloseq_tourist_clean1 <- phyloseq_tourist_clean %>%
  prune_taxa(taxa_sums(.) > 1, .)
phyloseq_tourist_clean1

#phyloseq-class experiment-level object
#otu_table()   OTU Table:         [ 1746 taxa and 135 samples ]
#sample_data() Sample Data:       [ 135 samples by 12 sample variables ]
#tax_table()   Taxonomy Table:    [ 1746 taxa by 7 taxonomic ranks ]
```

# Rarefaction 
There are contrasting opinions on rarefaction (to rarefy or not to rarefy) so, based on your question, it's up to you :)
Our suggestion is: if you want to compare microbiomes across samples --> RAREFY
                   if you want to describe microbiomes without any comparisons --> DO NOT RAREFY
Rarefaction is an essential step in order to compare samples across treatments or other variables of interest.
It is fundamental because without it we would compare samples without even reads. So, for example, we would compare samples with 1,000 reads
with others with 50,000 reads! That is no good!
Rarefaction avoids this problem
Drawback of rarefaction: possibility to lose samples and information
That's why is important to create a rarefaction curve to decide at what sampling depth would make sense to rarefy your samples.
```{r}
#Create a rarefaction curve:
#png("rarefaction_curve.png", width = 1000, height = 800, res = 150)
otu.rare = otu_table(phyloseq_tourist_clean1)
otu.rare = as.data.frame(t(otu.rare))
sample_names = rownames(otu.rare)
otu.rarecurve = rarecurve(otu.rare, step = 500, label = T, col = "antiquewhite4", xlab = "Sequencing Depth", ylab="Observed OTUs")
rarecurve(otu.rare, step = 500, label = TRUE, col = "antiquewhite4",
          xlab = "Sequencing Depth", ylab = "Observed OTUs",
          xaxt = "n") # suppress default x-axis

# Add vertical line at 20000
abline(v = 2000, col = "blue", lty = 2, lwd = 2)

# Custom x-axis with commas for large numbers
axis(side = 1, 
     at = axTicks(1), 
     labels = comma(axTicks(1)))

#dev.off()
```
There are several ways to decide how to rarefy based on your rarefaction curve:
1) rarefy at the lowest depth: will avoid losing samples but you may lose a lot of information:
rarefied <- rarefy_even_depth(phyloseq_tourist_clean1, rngseed=1, sample.size=min(sample_sums(phyloseq_tourist_clean1)), replace=F) 
set.seed(1)
2) custom choice of rarefaction: (to me is a better solution):
rarefied <- rarefy_even_depth(phyloseq_tourist_clean1, sample.size = 2000, replace = FALSE, rngseed = TRUE)
set.seed(TRUE)
for both methods we suggest to copy and paste your results below.

```{r}
rarefied <- rarefy_even_depth(phyloseq_tourist_clean1, sample.size = 2000, replace = FALSE, rngseed = TRUE)
set.seed(TRUE)
```
rarefied output: 466OTUs were removed because they are no longer 
                 present in any sample after random subsampling
                 
```{r}
# get number of OTUs before and after rarefying 
ntaxa(phyloseq_tourist_clean1)  # Before
ntaxa(rarefied) #after

```

# alpha diversity 
```{r}
# Calculate alpha diversity metrics with phyloseq function "estimate_richness"
richness_rarefied <- estimate_richness(rarefied)
head(richness_rarefied)
# Extract metadata from the phyloseq object
metadata_df <- as.data.frame(sample_data(rarefied))
# Merge richness data with metadata based on sample names
richness_with_metadata <- merge(richness_rarefied, metadata_df, by = "row.names")
# Assign meaningful column names (optional)
# colnames(richness_with_metadata)[0] <- "SampleID"
# Convert row names to a column and remove the original row names column
#richness_with_metadata$SampleID <- rownames(richness_with_metadata)
# Now, 'richness_with_metadata' contains both richness data and metadata
print(richness_with_metadata)
```
 Make a dataframe and csv file of the diversity indexes
```{r}
Alpha_diversity_df <- data.frame(richness_with_metadata)
write.csv(Alpha_diversity_df, file = "Alpha_diversity_rarefied.csv")
```

Normality test for alpha diversity 
```{r}
shapiro.test(richness_with_metadata$Observed) #not-normal distributed
shapiro.test(richness_with_metadata$Shannon) # not-normal distributed
shapiro.test(richness_with_metadata$Simpson) # not-normal distributed
shapiro.test(richness_with_metadata$Chao1)  # not-normal distributed
```
Stats summary of alpha diversity 
```{r}
names(richness_with_metadata)
```

```{r}
# Create a new dataset in which we put only the indices we want 
alpha_index_df_rarefied_all <- richness_with_metadata %>%
  select(Observed, Chao1, Shannon, Simpson, Year, Site) 

# Create an object in which we specify to calculate mean, median and standard deviation
alphastats <- list(
  mean = ~mean (.x, na.rm = TRUE),
  median = ~median(.x, na.rm = TRUE),
  sd = ~sd(.x, na.rm = TRUE)
)

# Create a summary of the Alpha diversity with mean, median and sd of our samples grouped by sampling time and treatment in this case 
stats_summary_alpha_diversity_all <- alpha_index_df_rarefied_all %>%
  group_by(Year, Site) %>%
  summarise(across(c(Observed, Chao1, Shannon, Simpson), alphastats) )
```

save summary
```{r}
write.csv(file = "Summary_of_stats_Alpha_diversity_all_samples.csv", stats_summary_alpha_diversity_all)
```


```{r}
# Clean and order 'Site' column once
alpha_index_df_rarefied_all <- alpha_index_df_rarefied_all %>%
  mutate(Site = gsub("_", " ", Site)) %>%
  mutate(Site = factor(Site, levels = c(
    "Directors Bay", "Playa Jeremy", "Holiday Beach",
    "Coral Estate", "Cas Abou", "Mambo Beach",
    "Porto Marie", "Grote Knip"
  )))
```

# Plots 
alpha diversity Shannon
```{r}
alpha_div_H <- alpha_index_df_rarefied_all %>%
  pivot_longer(cols = c(Shannon), names_to = "index", values_to = "value") %>%
  mutate(index_sorted = factor(index, levels = c("Shannon"), labels = c("OTUs"))) %>%
  ggplot(aes(x = Year, y = value, fill = Site)) +
  geom_boxplot(outlier.shape = NA) + 
  theme_classic() + theme(panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     axis.title = element_text(size=16),
                     axis.text.y = element_text(size = 14)) +
  labs(title = "H'",
       y = "Diversity index") +
#  geom_pwc(
#      aes(group = Year),
#      method = "dunn_test",
#      label = "p.signif",
#      p.adjust.method = "bonferroni",
#      hide.ns = TRUE,
#      tip.length = 0.03
#    ) +
  scale_fill_viridis_d(option = "C") +
  theme(legend.position = "none", strip.text = element_text(colour = "black"), strip.background = element_rect(fill = "white", colour = "black"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size = 14))
alpha_div_H
```

save Shannon plot 
```{r}
ggsave("alpha_diversity_Shannon.png", plot = alpha_div_H)
```

alpha diversity Chao1
```{r}
alpha_div_C1 <- alpha_index_df_rarefied_all %>%
  pivot_longer(cols = c(Chao1), names_to = "index", values_to = "value") %>%
  mutate(index_sorted = factor(index, levels = c("Chao1"), labels = c("OTUs"))) %>%
  ggplot(aes(x = Year, y = value, fill = Site)) +  ylim(0,200)+
  geom_boxplot(outlier.shape = NA) + 
  theme_classic() + theme(panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     axis.title = element_text(size=16),
                     axis.text.y = element_text(size = 14)) +
    labs(title = "Chao1",
             y = "Diversity index") +
 # geom_pwc(
 #     aes(group = Year),
 #     method = "dunn_test",
 #     label = "p.signif",
#    p.adjust.method = "bonferroni",
#      hide.ns = TRUE,
#      tip.length = 0.03
#    ) +
   scale_fill_viridis_d(option = "C") +
  theme(legend.position = "none", strip.text = element_text(colour = "black"), strip.background = element_rect(fill = "white", colour = "black"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size = 14))
alpha_div_C1
```

save Chao1 plot
```{r}
ggsave("alpha_diversity_chao1.png", plot = alpha_div_C1)
```

alpha diversity Simpson
```{r}
alpha_div_Simp <- alpha_index_df_rarefied_all %>%
  pivot_longer(cols = c(Simpson), names_to = "index", values_to = "value") %>%
  mutate(index_sorted = factor(index, levels = c("Simpson"), labels = c("OTUs"))) %>%
  ggplot(aes(x = Year, y = value, fill = Site)) +  ylim(0,1)+
  geom_boxplot(outlier.shape = NA) + 
  theme_classic() + theme(panel.grid.major = element_blank(),
                     panel.grid.minor = element_blank(),
                     axis.title = element_text(size=16),
                     axis.text.y = element_text(size = 14)) +
  labs(title = "Simpson",
       y = "Diversity index") +
 # geom_pwc(
#      aes(group = Year),
#      method = "dunn_test",
#      label = "p.signif",
#      p.adjust.method = "bonferroni",
#      hide.ns = TRUE,
#      tip.length = 0.03
#    ) +
   scale_fill_viridis_d(option = "C") +
  theme(legend.position = "none", strip.text = element_text(colour = "black"), strip.background = element_rect(fill = "white", colour = "black"),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(size = 14))
alpha_div_Simp
```

save Simpson plot
```{r}
ggsave("alpha_diversity_Simp.png", plot = alpha_div_Simp)
```

## PCoA
Conduct PCoA both for Year and Site
```{r}
png("PCoA_bac_years.png", width = 1000, height = 800, res = 150)

bray_rarefied.ord <- ordinate(rarefied, "PCoA", "bray")
ordination_scores <- as.data.frame(bray_rarefied.ord$vectors)
ordination_scores$Site <- sample_data(rarefied)$Site
colnames(ordination_scores)[1:2] <- c("PC1", "PC2")
centroids <- ordination_scores %>%
  group_by(Year) %>%
  summarise(PC1 = mean(PC1), PC2 = mean(PC2))
p_bray <- plot_ordination(rarefied, bray_rarefied.ord, type = "samples", color = "Year") + 
  theme_classic() 
p_bray <- p_bray + 
  geom_point(data = centroids, aes(x = PC1, y = PC2, color = Site), shape = 4, size = 4, stroke = 2) +
 scale_color_manual(values = c("2020" = "#440154", "2021" = "#f98e09", "2022" = "#21918c", "2023" = "#b73779"),
                     labels = c("2020", "2021", "2022", "2023")) +

p_bray

dev.off()
```

```{r}
png("PCoA_bac_sites.png", width = 1000, height = 800, res = 150)

highlight_site <- "Mambo_Beach" 

bray_rarefied.ord <- ordinate(rarefied, "PCoA", "bray")
ordination_scores <- as.data.frame(bray_rarefied.ord$vectors)
ordination_scores$Site <- sample_data(rarefied)$Site
colnames(ordination_scores)[1:2] <- c("PC1", "PC2")
centroids <- ordination_scores %>%
  group_by(Site) %>%
  summarise(PC1 = mean(PC1), PC2 = mean(PC2))
p_bray <- plot_ordination(rarefied, bray_rarefied.ord, type = "samples", color = "Site") + 
  theme_classic() 
p_bray <- p_bray + 
  geom_point(data = centroids, aes(x = PC1, y = PC2, color = Site), shape = 4, size = 4, stroke = 2) +
   stat_ellipse(data = subset(ordination_scores, Site != "Mambo_Beach"), 
               aes(x = PC1, y = PC2, group = Site), 
               alpha = 0.5) +
  stat_ellipse(data = subset(ordination_scores, Site == "Mambo_Beach"), 
               aes(x = PC1, y = PC2, group = Site), 
                size = 1)

p_bray

dev.off()
```

## PERMANOVA
```{r}
rarefied_transformed <- rarefied %>%
  transform_sample_counts(function(x) 100* {x/sum(x)})

phyloseq1_bray <- phyloseq::distance(rarefied_transformed, method = "bray")
bray_dist <- as.matrix(dist(phyloseq1_bray))
sampledf <- data.frame(sample_data(rarefied_transformed))
adonis_bray_year <- adonis2(phyloseq1_bray ~ Year, data = sampledf)
adonis_bray_year
adonis_bray_site <- adonis2(phyloseq1_bray ~ Site, data = sampledf)
adonis_bray_site
```

check assumption 
```{r}
beta_species <- betadisper(phyloseq1_bray, sampledf$Site)
permutest(beta_species)
```

```{r}
beta_samplingtime <- betadisper(phyloseq1_bray, sampledf$Year)
permutest(beta_samplingtime)
```

## Taxonomy 
Set taxonomic rank
```{r}
genusabundance <- rarefied %>%
  tax_glom(taxrank = "Genus") %>%                    # Set to smallest taxonomic level you are interested in
  transform_sample_counts(function(x) {x/sum(x)} ) %>%   # Transform to rel. abundance
  psmelt()                                               # Melt to long format

```

```{r}
genusabundance <- genusabundance %>%
  mutate(Site = gsub("_", " ", Site)) %>%
  mutate(newLabel = paste(Status, Site, sep = "-"))
head(genusabundance)
```

filter genus abundance dataframe to show only desired taxonmic levels
```{r}
##entire dataset
all <- genusabundance %>%
  select(Phylum, Class, Family, Genus, Sample, Abundance,Year, Site, Label, newLabel) %>%
  filter(Abundance != 0) %>%
  mutate(
    Phylum = as.character(Phylum),
    Class = as.character(Class),
    Family = as.character(Family),
    Genus = as.character(Genus))
head(all)

```

calculate rel. and total abundances and group in >5% if applicabale 
```{r}

## Group data and calculate total and relative abundances
relative_abundance_family <- all %>%
  group_by(Family, Year, Site, Label, newLabel) %>%
  summarise(TotalAbundance = sum(Abundance), .groups = "drop") %>%
  group_by(Label) %>%
  mutate(RelativeAbundance = TotalAbundance / sum(TotalAbundance))

#all families that are present in only less than 5% samples are grouped in one group "< 5%"
relative_abundance_family_morethan5 <- relative_abundance_family %>%
  mutate(Family = ifelse(RelativeAbundance < 0.05, "< 5 %", Family))

relative_abundance_family_morethan5 <- relative_abundance_family_morethan5 %>%
  group_by(Family, Year, Site, Label, newLabel) %>%
  summarise(RelativeAbundance = sum(RelativeAbundance), .groups = "drop")
```

```{r}
library(colorspace)
desired_order <- c(
    "CONTROL-Directors Bay", "CONTROL-Playa Jeremy", "URBAN-Holiday Beach",
    "URBAN-Coral Estate", "URBAN-Cas Abou", "TOURIST-Mambo Beach",
    "TOURIST-Porto Marie", "TOURIST-Grote Knip"
  )

# Create a mapping between newLabel and Label
label_map <- relative_abundance_family_morethan5 %>%
  select(Label, newLabel) %>%
  distinct() %>%
  filter(newLabel %in% desired_order)

# Create a factor with levels ordered by desired newLabel
label_ordered <- label_map %>%
  arrange(factor(newLabel, levels = desired_order)) %>%
  pull(Label)

# Apply the new ordering to Label
relative_abundance_family_morethan5$Label <- factor(
  relative_abundance_family_morethan5$Label,
  levels = label_ordered
)
```


stacked bar plot of abundance per sample per years
```{r}
#create a custom color palette 
colorBlindness::displayColors(SteppedSequential5Steps) 
color_palette <- SteppedSequential5Steps 
# Name the palette using unique Family levels
names(color_palette) <- unique(relative_abundance_family_morethan5$Family)

# Override the color
color_palette["< 5 %"] <- "grey"
color_palette["SAR86_clade_fa"] <- "#51A3CC"


#create plot 
relative_abundance_plot <- ggplot(relative_abundance_family_morethan5, aes(x = Label, y = RelativeAbundance, fill = Family)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  labs(x = "Sample site",
       fill = "Family") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  facet_wrap(~ Year, scales = "free_y", nrow = 4) + 
  scale_fill_manual(values = color_palette)  +
  scale_x_discrete(breaks = relative_abundance_family_morethan5$Label,
                     labels = relative_abundance_family_morethan5$newLabel) +
  coord_flip() +
  theme(legend.position = "right")
  

print(relative_abundance_plot)

ggsave("bac_stackedbarplot_horizontal.pdf", plot = relative_abundance_plot,
       width = 20,
       height = 10,
       units = "in")
```

#ANCOMBC
prepare subsets of years that for comparison and create phyloseqs
```{r}
 year_pairs <- list(c("2020", "2021"),
                   c("2021", "2022"),
                   c("2022", "2023"))
#Years 2020 and 2021
phy_20_21 <- subset_samples(phyloseq_tourist_clean1, Year %in% c("2020", "2021"))
phy_20_21 <- prune_taxa(taxa_sums(phy_20_21) > 0, phy_20_21)
#Years 2021 and 2022
phy_21_22 <- subset_samples(phyloseq_tourist_clean1, Year %in% c("2021", "2022"))
phy_21_22 <- prune_taxa(taxa_sums(phy_21_22) > 0, phy_21_22)
#Years 2022 and 2023
phy_22_23 <- subset_samples(phyloseq_tourist_clean1, Year %in% c("2022", "2023"))
phy_22_23 <- prune_taxa(taxa_sums(phy_22_23) > 0, phy_22_23)
```

 run ANCOMBC on phyloseqs
```{r}
set.seed(124)

## 2020 2021
ancom_20_21 <- ancombc2(phy_20_21, fix_formula = "Year", p_adj_method = "BH", lib_cut = 1000, group = "Year", struc_zero = FALSE, neg_lb = FALSE, alpha = 0.05, global = FALSE 
) 

res_20_21 <- ancom_20_21$res
sum(res_20_21$diff_Year.L == TRUE) #83 significant OTUs
nrow(res_20_21) # 215 OTUs analyzed

save(ancom_20_21, file = "ancombc_20_21_c.RData")

##2021 2022
ancom_21_22 <- ancombc2(phy_21_22, fix_formula = "Year", p_adj_method = "BH", lib_cut = 1000, group = "Year", struc_zero = FALSE, neg_lb = FALSE, alpha = 0.05, global = FALSE 
) 

res_21_22 <- ancom_21_22$res
sum(res_21_22$diff_Year.L == TRUE) #127 significant OTUs
nrow(res_21_22) #259 OTUs analyzed

save(ancom_20_21, file = "ancombc_21_22_c.RData")

##2022 2023
ancom_22_23 <- ancombc2(phy_22_23, fix_formula = "Year", p_adj_method = "BH", lib_cut = 1000, group = "Year", struc_zero = FALSE, neg_lb = FALSE, alpha = 0.05, global = FALSE 
) 

res_22_23 <- ancom_22_23$res
sum(res_22_23$diff_Year.L == TRUE) #155 significant OTUs
nrow(res_22_23)  #279 OTUs analyzed
save(ancom_22_23, file = "ancombc_22_23_c.RData")
```

Prepare data
```{r}
test_tax_20_21 <- tax_table(phy_20_21)
dimnames(test_tax_20_21)
tax_table_gen_20_21 <- as_tibble(test_tax_20_21@.Data) %>%
  mutate(taxon = dimnames(test_tax_20_21)[[1]]) %>%  
  relocate(taxon, .before = 1)

test_tax_21_22 <- tax_table(phy_21_22)
dimnames(test_tax_21_22)
tax_table_gen_21_22 <- as_tibble(test_tax_21_22@.Data) %>%
  mutate(taxon = dimnames(test_tax_21_22)[[1]]) %>%  
  relocate(taxon, .before = 1)

test_tax_22_23 <- tax_table(phy_22_23)
dimnames(test_tax_22_23)
tax_table_gen_22_23 <- as_tibble(test_tax_22_23@.Data) %>%
  mutate(taxon = dimnames(test_tax_22_23)[[1]]) %>%  
  relocate(taxon, .before = 1)

```

Ancombc results table alpha 0.05
```{r}
ancom_gg_data_20_21 <- res_20_21 %>%
  left_join(tax_table_gen_20_21[,c("taxon", "Order", "Family", "Genus")]) %>%
  filter(diff_Year.L == TRUE) %>%
  mutate(wg = ifelse(W_Year.L>0, "pos", "neg")) %>%
  mutate(new_lab = paste(taxon, Genus))  

ancom_gg_data_df_20_21 <- data.frame(ancom_gg_data_20_21, nrow=nrow(ancom_gg_data_20_21))

ancom_gg_data_21_22 <- res_21_22 %>%
  left_join(tax_table_gen_21_22[,c("taxon", "Order", "Family", "Genus")]) %>%
  filter(diff_Year.L == TRUE) %>%
  mutate(wg = ifelse(W_Year.L>0, "pos", "neg")) %>%
  mutate(new_lab = paste(taxon, Genus))  

ancom_gg_data_df_21_22 <- data.frame(ancom_gg_data_21_22, nrow=nrow(ancom_gg_data_21_22))

ancom_gg_data_22_23 <- res_22_23 %>%
  left_join(tax_table_gen_22_23[,c("taxon", "Order", "Family", "Genus")]) %>%
  filter(diff_Year.L == TRUE) %>%
  mutate(wg = ifelse(W_Year.L>0, "pos", "neg")) %>%
  mutate(new_lab = paste(taxon, Genus))  

ancom_gg_data_df_22_23 <- data.frame(ancom_gg_data_22_23, nrow=nrow(ancom_gg_data_22_23))
```

Number of increased and decreased OTUs
```{r}
ancom_gg_data_20_21 %>%
  count(wg)

ancom_gg_data_21_22 %>%
  count(wg)

ancom_gg_data_22_23 %>%
  count(wg)

```

Volcano plots
```{r}
process_ancom_pair <- function(ancom_result, phyloseq_obj, comparison_label) {
  # Extract results
  res <- ancom_result$res
  
  # Get taxonomy
  tax_table_df <- as.data.frame(tax_table(phyloseq_obj)) %>%
    rownames_to_column("taxon")
  
  # Transform to relative abundance
  rel_abund_matrix <- vegan::decostand(otu_table(phyloseq_obj), method = "total", MARGIN = 2)
  rel_abund_df <- data.frame(relmean = rowMeans(rel_abund_matrix)) %>%
    rownames_to_column("taxon")
  
  # Merge all
  volcano_df <- res %>%
    select(taxon, starts_with("W_"), starts_with("lfc_"), starts_with("q_"), starts_with("diff_")) %>%
    left_join(tax_table_df, by = "taxon") %>%
    left_join(rel_abund_df, by = "taxon") %>%
    mutate(
      group = ifelse(lfc_Year.L > 0, "Enriched", "Decreased"),
      group2 = ifelse(diff_Year.L == FALSE, "Not significant", group),
      group2 = factor(group2, levels = c("Decreased", "Enriched", "Not significant")),
      sizept = log2(relmean) + 15,
      comparison = comparison_label
    )
  
  return(volcano_df)
}

```


```{r}
volcano_20_21 <- process_ancom_pair(ancom_20_21, phy_20_21, "2020 vs 2021")
volcano_21_22 <- process_ancom_pair(ancom_21_22, phy_21_22, "2021 vs 2022")
volcano_22_23 <- process_ancom_pair(ancom_22_23, phy_22_23, "2022 vs 2023")

# Combine for plotting
all_volcano_data <- bind_rows(volcano_20_21, volcano_21_22, volcano_22_23)

```

```{r}
ggplot(all_volcano_data, aes(x = lfc_Year.L, y = -log10(q_Year.L), color = group2)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ comparison, scales = "free") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(log2(0.9), log2(1.1)), linetype = "dashed") +
  scale_color_manual(values = c("brown", "darkcyan", "gray"), name = "OTUs") +
  labs(x = "Log Fold Change", y = "-log10(q-value)", title = "Volcano Plot for Year Comparisons") +
  theme_classic()

```

Create new color palette that is in accordance with the relative abundance plot
```{r}

all_families <- unique(top_otus_by_comparison$Family)

less5_label <- "< 5 %"

grey_families <- names(color_palette)[color_palette == "grey" & names(color_palette) != less5_label]


n_soft <- length(grey_families)

if (n_soft > 0) {
  soft_colors <- brewer.pal(min(8, n_soft), "Pastel1")
  if (n_soft > length(soft_colors)) {
    soft_colors <- colorRampPalette(soft_colors)(n_soft)
  }
  
  soft_family_colors <- setNames(soft_colors, grey_families)
} else {
  soft_family_colors <- c()
}

final_palette <- color_palette

final_palette[grey_families] <- soft_family_colors

final_palette[less5_label] <- "grey"

```

Create top 20 table across comparisons and create plots 
```{r}
top_otus_by_comparison <- all_volcano_data %>%
  filter(diff_Year.L == TRUE, Genus != "Unclassified") %>%
  group_by(comparison) %>%
  slice_max(order_by = abs(W_Year.L), n = 20) %>%
  ungroup() %>%
  mutate(taxon_label = paste(taxon , Genus, sep = " "))
         
top_otus_by_comparison <- top_otus_by_comparison %>%
  mutate(Family = factor(Family, levels = names(final_palette)),
         # create taxon_label combining taxon, Family, Genus separated by spaces
         taxon_label = paste(taxon, Genus, sep = " "))

plot_top_otus <- function(data, comparison_name) {
  plot_data <- data %>%
    filter(comparison == comparison_name) %>%
    mutate(taxon_label = factor(taxon_label, levels = taxon_label[order(W_Year.L)]))
  
  ggplot(plot_data, aes(x = taxon_label, y = W_Year.L, fill = Family)) +
    geom_col(color = "black", linewidth = 0.4, alpha = 0.95) +
    coord_flip() +
      scale_fill_manual(values = final_palette) +
    labs(
      title = paste("Top 20 Significant OTUs:", comparison_name),
      x = "OTU (Family + Genus)",
      y = "W Statistic"
    ) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5),
     axis.text.y = element_text(colour = "black"),
      legend.position = "bottom"
    )
}
```


Save individual plots
```{r}
p1 <- plot_top_otus(top_otus_by_comparison, "2020 vs 2021")
p2 <- plot_top_otus(top_otus_by_comparison, "2021 vs 2022")
p3 <- plot_top_otus(top_otus_by_comparison, "2022 vs 2023")

ggsave("top20_otus_2020_vs_2021.pdf", plot = p1, width = 10, height = 6)
ggsave("top20_otus_2021_vs_2022.pdf", plot = p2, width = 10, height = 6)
ggsave("top20_otus_2022_vs_2023.pdf", plot = p3, width = 10, height = 6)
```
